---
title: "Polynomial Library in BsplineQuantReg"
author: "Alexandre Abbes"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Polynomial Library}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 7, fig.height = 5)
library(BsplineQuantReg)
```

## Introduction

The `BsplineQuantReg` package provides a comprehensive set of tools for polynomial manipulation, implemented in pure R with a consistent interface. All polynomial functions follow the **decreasing power order convention**, where a polynomial $p(x) = a_0 + a_1 x + a_2 x^2$ is represented as `c(a_2, a_1, a_0)`.

This vignette covers: - Basic polynomial operations (addition, multiplication, evaluation, differentiation) - Piecewise polynomial (PP) form - Callable and non-callable PP objects - Extracting parameters from PP objects

## Basic Polynomial Operations

### Polynomial Multiplication

```{r polymul}
# (1 + x) * (1 + x) = 1 + 2x + x^2
p1 <- c(1, 1)  # 1 + x
p2 <- c(1, 1)  # 1 + x
product <- polymul(p1, p2)
print(product)  # c(1, 2, 1) → 1 + 2x + x^2

# (x^2 + 2x + 1) * (x - 1) = x^3 + x^2 - x - 1
p3 <- c(1, 2, 1)  # x^2 + 2x + 1
p4 <- c(1, -1)    # x - 1
polymul(p3, p4)
```

### Polynomial Addition

```{r polyadd}
# (1 + x) + (1 - x) = 2
polyadd(c(1, 1), c(1, -1))

# (x^2 + 1) + (x + 1) = x^2 + x + 2
polyadd(c(1, 0, 1), c(1, 1))
```

### Polynomial Evaluation

```{r poly_eval}
# P(x) = 1 + x + x^2
p <- c(1, 1, 1)  # x^2 + x + 1
poly_eval(p, c(0, 1, 2))  # returns c(1, 3, 7)

# Evaluate at many points
x <- seq(-2, 2, length.out = 10)
y <- poly_eval(c(1, 0, -1), x)  # 1 - x^2
plot(x, y, type = "l", main = "1 - x^2")
```

### Polynomial Differentiation

```{r polyderiv}
# P(x) = x^2 → P'(x) = 2x
p <- c(1, 0, 0)  # x^2
polyderiv(p, der = 1)  # c(2, 0) → 2x

# Second derivative: P''(x) = 2
polyderiv(p, der = 2)  # c(2)

# Higher order derivatives
p <- c(1, 2, 3, 4)  # 4x^3 + 3x^2 + 2x + 1
polyderiv(p, der = 2)  # 24x + 6
```

### Polynomial Reduction

```{r reduce_pol}
# Remove leading zeros
reduce_pol(c(0, 0, 1, 2, 1))  # c(1, 2, 1) → x^2 + 2x + 1
reduce_pol(c(0, 0, 0, 5))     # c(5) → constant polynomial
```

### Taylor Basis Change

```{r change_polynomial_base_taylor}
# Convert (x-1)^2 to expansion around 0
# (x-1)^2 = x^2 - 2x + 1
coeffs_a <- c(1, 0, 0)  # (x-1)^2 in basis centered at a=1
change_polynomial_base_taylor(coeffs_a, a = 1, b = 0)
# Returns c(1, -2, 1) → x^2 - 2x + 1
```

## Piecewise Polynomial (PP) Form

The PP form represents a function defined piecewise by polynomials on different intervals.

### Creating a PP Object

```{r makpp}
# Create a piecewise polynomial with two intervals
# Interval 1: x^2 on [0, 1]
# Interval 2: 2x - 1 on [1, 2]
coeff <- matrix(c(
  1, 0, 0,   # x^2
  0, 2, -1   # 2x - 1
), nrow = 2, byrow = TRUE)

knots <- c(0, 1, 2)

pp <- makpp(coeff, knots)
print(pp)

# Evaluate the PP form
x <- seq(0, 2, length.out = 100)
y <- evalpp(pp, x)

plot(x, y, type = "l", main = "Piecewise Polynomial")
abline(v = knots, col = "red", lty = 2)
```

### Non-Callable PP Objects

```{r non_callable_pp}
# A non-callable PP object is a list with components:
# - coeff: matrix of polynomial coefficients
# - knot: knot positions
# - degree: polynomial degree

pp <- makpp(coeff, knots, callable = FALSE)
class(pp)  # "non_callable_pp"

# Access components
pp$coeff
pp$knot
pp$degree
```

### Callable PP Objects

A callable PP object can be evaluated directly like a function.

```{r callable_pp}
# Create a callable PP
pp_call <- makpp(coeff, knots, callable = TRUE)
class(pp_call)  # "callable_pp" "function"

# Evaluate directly
x <- seq(0, 2, length.out = 10)
y <- pp_call(x)

# Print shows information
print(pp_call)

# Can be used in plots
plot(pp_call, xlim = c(0, 2))
```

### Extrapolating values

When evalpp() is called with x values outside the knot range (x \< knots[1] or x \> knots[length(knots)]), it automatically performs extrapolation by extension using the border polynomial piece (the first or last polynomial in the PP structure). For example, if the knot vector is c(0, 1, 2), points with x \< 0 will be evaluated using the polynomial from the first interval [0, 1], while points with x \> 2 will use the polynomial from the last interval [1, 2]. This is equivalent to "extending" the first and last piece of the spline. A warning message is printed to alert the user when extrapolation occurs.

```{r}
# PP with two intervals: x^2 on [0,1], 2x-1 on [1,2]
coeff <- matrix(c(1, 0, 0, 0, 2, -1), nrow = 2, byrow = TRUE)
knots <- c(0, 1, 2)
pp <- makpp(coeff, knots, callable = TRUE)

# Evaluate inside and outside the knot range
x_test <- c(-0.5, 0.5, 1.5, 2.5)
y_test <- pp(x_test)
data.frame(x = x_test, y = y_test)
# The values at -0.5 and 2.5 are extrapolated
```

### Extracting Parameters

```{r get_parameters}
# For callable PP objects, use get_parameters()
params <- get_parameters(pp_call)
print(params$degree)
print(params$knot)
print(params$coeff)

# For non-callable PP, access directly
pp=makpp(pp,callable=FALSE,verbose=TRUE)
pp$degree
pp$knot
pp$coeff
```

## Advanced: PP Form Manipulation

### Extracting a Single Interval Polynomial

```{r extract_interval}
# Get polynomial for interval i
pp=makpp(pp,callable=FALSE)
interval_poly <- pp$coeff[1, ]  # First interval
print(interval_poly)

# Evaluate on a subinterval
x_sub <- seq(0, 1, length.out = 20)
y_sub <- poly_eval(interval_poly, x_sub)
```

### Combining PP Objects

```{r combine_pp}
# Create two PP objects and combine them
# This is useful for constructing complex piecewise functions
coeff1 <- matrix(c(1, 0), nrow = 1)
coeff2 <- matrix(c(0, 1), nrow = 1)

pp1 <- makpp(coeff1, c(0, 1))
pp2 <- makpp(coeff2, c(1, 2))

# Combine coefficients and knots
combined_coeff <- rbind(coeff1, coeff2)
combined_knots <- c(0, 1, 2)
pp_combined <- makpp(combined_coeff, combined_knots)

# Evaluate combined function
x <- seq(0, 2, length.out = 100)
y_combined <- evalpp(pp_combined, x)
plot(x, y_combined, type = "l")
abline(v = c(0, 1, 2), col = "red", lty = 2)
```

### Visualizing PP with Basis Functions

```{r visualize_pp}
# Create a PP with multiple pieces
knots <- seq(0, 1, length.out = 5)
degree <- 2
n_pieces <- length(knots) - 1

# Generate random coefficients
coeff <- matrix(rnorm(n_pieces * (degree + 1)), nrow = n_pieces)

pp <- makpp(coeff, knots, callable = TRUE)

# Evaluate and plot
x <- seq(0, 1, length.out = 200)
y <- pp(x)

plot(x, y, type = "l", lwd = 2,
     main = "Random Piecewise Quadratic")
abline(v = knots, col = "red", lty = 2)
grid()
```

## Comparison: PP vs B-spline

```{r pp_vs_bspline}
# Create a B-spline and convert to PP form
sn <- c(0, 0, 0, 0, 0.25, 0.5, 0.75, 1, 1, 1, 1)
basis <- Bspline_base(sn, degree = 3)
basis$coeff <- runif(basis$n_splines)

# Convert to PP
pp_from_bspline <- Bsplinetopp(basis)

# Both represent the same function
x <- seq(0, 1, length.out = 100)
y_bspline <- spline_eval(basis, x)
y_pp <- evalpp(pp_from_bspline, x)

# They should match
max(abs(y_bspline - y_pp))
```

## Summary

<!-- | Feature | Function | Description | -->

<!-- |---------|----------|-------------| -->

<!-- | Polynomial multiplication | `polymul()` | Multiply two polynomials | -->

<!-- | Polynomial addition | `polyadd()` | Add two polynomials | -->

<!-- | Polynomial evaluation | `poly_eval()` | Evaluate at points | -->

<!-- | Polynomial derivative | `polyderiv()` | Differentiate polynomial | -->

<!-- | Taylor shift | `change_polynomial_base_taylor()` | Change expansion point | -->

<!-- | PP creation | `makpp()` | Create PP form | -->

<!-- | PP evaluation | `evalpp()` | Evaluate PP at points | -->

<!-- | Callable PP | `makpp(callable = TRUE)` | PP as function | -->

<!-- | Extract params | `get_parameters()` | Extract from callable | -->

<!-- All functions are implemented in pure R and follow matlab consistent conventions, making them robust and easy to use for B-spline manipulation and quantile regression. -->
